Density-dependent selection and the limits of relative fitness

### 2017-12-08 manuscript.tex
l

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$,  $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype $j$ and a mutant $i$ for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density and cannot be treated as a simple constant (the possible frequency-dependence of $s$ is a separate issue, because this still leaves a closed model in terms of the frequencies). This issue is evaded by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of the population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope). The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp. 468]{barton_2007} \citep{prout_1980}. While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can depend on density. In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.

@@

Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation''). We then...

@@

We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

@@

The number of individuals of the $i$'th type landing in any particular territory is denoted $x_i$. We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density. This is approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).

@@

\subsection*{Mean-field approximation of the density-dependent lottery}

Eq. \eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition. We now evaluate this expectation using a ``mean field'' approximation.

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of mean field approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density Fig.~\ref{fig:LVvslottery}b. Selection then occurs purely along this line, uncoupled from the density regulation of $N$. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. This is all perfectly consistent with MacArthur's general argument.

\subsection*{Density-dependent selection and the strength of selection}

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even in crowded populations where $N$ has reached equilibrium, because adaptive sweeps in density-dependent traits then induce a change in $N$ and also change themselves in response to $N$. To quantify how serious this objection is, we need to account for how much $N$ changes over a sweep. In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N=b_i/\delta_i$, and so the selection coefficient increases from $s_{\rm initial}$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a sweep (Fig.~\ref{fig:strengthofselection}). Thus, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on the density-dependent mortality rate, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.

@@

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:Ndependentsweep} with our density-dependent lottery. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$. Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits. And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.

@@

It is widely recognized that the evolutionary description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. Typically this is cast as a narrow focus on frequencies on the part of evolutionary geneticists, who, perhaps out of necessity, omit the complexities of population ecology. While there may be some truth to this, we argue that part of the reason for the disconnect between the evolutionary geneticist's selection and the ecologist's selection is that relative fitness models actually contain important ecological processes omitted in many ecological models, including the classical Lotka-Volterra and logistic models. To illustrate, we highlight two important properties of our extended lottery model.

First, our lottery model contains a ``reproductive excess'' [any good suggestions for citations here, Joanna?] that allows selection (and competition) to occur among juveniles without affecting population density at the adult stage. This excess appears when the number of propagules produced is greater than the number of available territories. By contrast, in classical ecological growth models only one life-history stage is represented and selection is intrinsically connected to the effects of each type's adult density on the other. These are restrictive assumptions --- it is often the case that competition and selection are concentrated at the more fragile juvenile stage \citep{chesson_1983}. Reproductive excesses tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of the large amount of selection occurring at the juvenile stage. Reproductive excesses expand the domain of validity of Eq.~\eqref{eq:canonical} by facilitating selection-independent density \citep{prout_1980}.

Second, density is regulated by fecundity $b$ and mortality $d$ traits that are affected equally by density. There are only so many territories, and all territories are equally available to all types. This removes the density-dependence that would make selection on $b$ and $d$ exhibit the difficulties shown in Fig.~\ref{fig:strengthofselection} without removing density-dependent selection from the model as whole via $c$ (by contrast, \cite[pp. 468]{barton_2007} allows density to depend on selection but removes density-dependent selection entirely). Although standard ecological models do have type-specific regulation of density (setting aside special parameter choices \citep{mallet_2012}), it is not obvious that we should expect this as a general rule. Much of our intuition about type-specific regulation of density (and density-dependent selection in general) is grounded in the logistic model and its diploid bi-allelic variants \citep{roughgarden_1979}, which should hardly be regarded as universal \citep{christiansen_2004,mallet_2012}. Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition. Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$. Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.

Relative fitness breaks down if strong selection occurs in a way that changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In our lottery model, the latter can only occur if types differ in $c$ as well as $b$ or $d$. The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.

>>>>

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection (we would also need to specify a model for density). Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Counter to the assumption of constant $N$, MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection versus relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope). The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp. 468]{barton_2007} \citep{prout_1980}. While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can, in reality, depend on density. In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

@@

Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...

@@

We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance. We derive equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

@@

The number of individuals of the $i$'th type landing in any particular territory is denoted $x_i$. We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density. This approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).

@@

\subsection*{Analytical approximation of the density-dependent lottery}

Eq. \eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition. We now evaluate this expectation.

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of the analytical approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types (Fig.~\ref{fig:LVvslottery}a).

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical (and hence so is saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b). Selection then occurs purely along this line, uncoupled from the regulation of density. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.

\subsection*{Density-dependence and the strength of selection}

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even when $N$ has reached equilibrium. To quantify how serious this objection is, we need to account for how much $N$ changes as a result of a sweep involving a density-regulating trait. In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep. Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep (Fig.~\ref{fig:strengthofselection}). Consequently, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on $\delta_i$, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection. The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.

@@

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly constant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$. Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits. And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly. [Need to flesh this out a little more and address some subtleties with discrete vs. continuous time]

@@

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, unfairly invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

Having defended Eq.~\eqref{eq:canonical}

, relative fitness does indeed break down if strong selection occurs in a way that changes (adult) population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In our lottery model, the this can only occur if types differ in $c$ as well as $b$ or $d$. The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate. This may seem odd given the behavior of most models in the density-dependent literature, so it worth investigating this point further.

The essential distinction between selection in uncrowded and crowded conditions is the absence or presence of interactions between types (these interactions may be indirect, as in the case of consumable resource competition). That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: (i) some types expand more rapidly in the absence of interactions between types or (ii) some types are superior in their interactions with other types. This distinction captures a key property of crowding, namely that you must displace others to grow.

By contrast, the separation of traits in the $r$/$K$ scheme is presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g. \citep{gill_1974}). As we have seen, MacArthur's general argument \citep{macarthur_1967} does not justify this dichotomy, because selection in crowded populations also involves inter-type competition represented by the Lotka-Volterra $\alpha_{ij}$ coefficients; selection on the latter has been termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. In the Lotka-Volterra model, $r_i$ belongs to (i) and the inter-type coefficients $\alpha_{ij}$ ($i\neq j$) belong to (ii). But $\alpha_{ii}$, the inverse of type $i$'s carrying capacity $K_{ii}$, straddles both (i) and (ii). For this reason, $K$-selection --- selection for greater carrying capacity --- is a particularly confusing way to characterize selection in crowded populations.

In our lottery model, $b$ and $d$ are interaction-independent traits, whereas $c$ is entirely about interactions between types.

(factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a)

(factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).

\citep{chesson_2000}

Another common feature of density-regulated growth models is the presence of traits which both regulate density and are subject to density-dependent selection, such as the $\delta$ trait in the simple birth-death model Eq.~\eqref{eq:simplebirthdeath}. The $b$/$\delta$ parameterization in Eq.~\eqref{eq:simplebirthdeath} emphasizes the obvious point that traits which control density need not be subject to density-dependent selection. The $K$ parameter in the standard $r/K$ parameterization of the logistic model has been particularly misleading in this respect, because it is solely responsible for the regulation of density and is subject to density-dependent selection. Nevertheless, $\delta$-like traits arise automatically in the simplest linear models of density-regulated growth, including the logistic and Lotka-Volterra models, regardless of how they are parameterized, and are thus deeply ingrained in the literature on density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012}. In our lottery model, $b$ and $d$ control density, but do not exhibit density-dependent selection, and thus do not pose the same threat to Eq.~\eqref{eq:canonical} as a $\delta$-like trait.
